---
title: "New BCI kmeans-gap analysis"
author: "Rafael D'Andrea"
date: '2022-06-17'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(SWKM)
library(furrr)
library(parallel)
library(rstatix)

theme_set(theme_bw())
theme_update(
  panel.grid = element_blank(),
  strip.background = element_rect(fill = 'orange')
)

cores = detectCores() - 1
plan(multisession, workers = cores)
```

```{r}
remove.outliers = TRUE
```


Census data from `G:/My Drive/Stony Brook University/Research/BCI/bci.tree/bci.tree7.rdata`.

Trait data from `G:/My Drive/Stony Brook University/Research/BCI/BCITRAITS_NEW.csv`.

Trees are counted which exceed half of the max dbh for their species.

Traits:

- height = HEIGHT_AVD
- leaf = log(LMALAM_AVD)
- seed = log(SEED_DRY)
- wood = SG60C_AVG

Number of nulls: 500

Metric: abundance-weighted k-means from package `SWKM`.

```{r}
bci = 
  get(load('G:/My Drive/Stony Brook University/Research/BCI/bci.tree/bci.tree7.rdata')) |> 
  as_tibble() |>
  drop_na(dbh)

zo_scale = function(x){
  max_x = max(x, na.rm = TRUE)
  min_x = min(x, na.rm = TRUE)
  return((x - min_x) / (max_x - min_x))
} 

traits = 
  read.csv('G:/My Drive/Stony Brook University/Research/BCI/BCITRAITS_NEW.csv') |> 
  as_tibble() |>
  mutate(sp = tolower(SP.)) |>
  select(
    sp, 
    genus = `GENUS.`, 
    species = `SPECIES.`, 
    height = HEIGHT_AVG, 
    wood = SG60C_AVG,
    seed = SEED_DRY,
    leaf = LMALAM_AVD,
    maxdbh = DBH_AVG
  ) |>
  mutate(
    seed = log(seed),
    leaf = log(leaf)
  )

dtf = 
  bci |>
  full_join(traits, by = 'sp')

abuns = 
  dtf |>
  group_by(sp) |>
  summarize(abun = sum(dbh >= maxdbh / 2)) |>
  drop_na()

dat = 
  abuns |>
  inner_join(traits, by = 'sp') |>
  mutate(
    across(height:leaf, zo_scale)
  ) |>
  pivot_longer(height:leaf, names_to = 'trait') |>
  drop_na()
  
```


```{r}
plot_boxes =
  dat |> 
  ggplot(aes(trait, value, fill = trait)) + 
  geom_boxplot() + 
  theme(legend.position = 'none')

plot_boxes |>
  show()
```


```{r}
if(remove.outliers){
  
  writeLines('Outliers are being removed from kmeans analysis')
  
  dat = 
    dat |>
    anti_join(
      dat |> 
        group_by(trait) |> 
        identify_outliers(value)
    )
}   
```


```{r}
plot = 
  dat |>
  ggplot() +
  geom_segment(
    aes(
      x = value, 
      y = rep(0, nrow(dat)), 
      xend = value, 
      yend = abun
      )
    ) +
  labs(x = 'Trait', y ='Abundance (dbh > dbh_max/2)') +
  coord_cartesian(ylim = c(0, 700)) +
  facet_wrap(~trait, scales = 'free')

plot |>
  show()

plot_hists = 
  dat |>
  ggplot(aes(value)) +
  geom_histogram() +
  facet_wrap(~trait, scales = 'free')
```


```{r eval=FALSE}

cents = matrix(quantile(dat$height, 1:4/4), 4, 1)

res = 
  kmeans.weight(
    x = matrix(dat$height, nrow(dat), 1), 
    K = 4, 
    nstart = 100,
    weight = dat$abun, centers = cents
  )

dat =
  dat |>
  mutate(cluster = as.factor(res$cluster))

plot1 = 
  dat |>
  ggplot() +
  geom_segment(
    aes(
      x = height, 
      y = rep(0, nrow(dat)), 
      xend = height, 
      yend = abun, 
      color = cluster
      )
    ) +
  labs(x = 'Species Trait', y ='Species Abundance') +
  coord_cartesian(ylim = c(0, 700)) +
  theme(legend.position = c(.9,.6))

plot1 |>
  show()

dat |>
  select(-maxdbh) |>
  write.csv('G:/My Drive/Stony Brook University/Research/BCI/BCI_heightclusters.csv')
```

```{r}

maxclust = 10

null_dtf =
  expand_grid(
    clusters = 2:maxclust,
    null = 0:500,
    traits = unique(dat$trait)
  )

myfun = 
  function(clusters, null, traits){
    
    data = 
      dat |>
      filter(trait == traits)
    
    set.seed(null)
    
    if (null == 0)
      ns = data$abun
    else
      ns = sample(data$abun)
    
    cents = matrix(quantile(data$value, 1:clusters / clusters), clusters, 1)
    
    wcss =
      kmeans.weight(
        x = matrix(data$value, nrow(data), 1),
        K = clusters,
        nstart = 100,
        weight = ns,
        centers = cents
      ) |>
      pluck('wcss')
    
    return(tibble(clusters = clusters, null = null, trait = traits, wcss = wcss))
  }

kmeanres = 
  null_dtf |>
  future_pmap_dfr(
    .f = myfun,
    .options = furrr_options(seed = NULL)
  )

stats = 
  kmeanres |> 
  group_by(trait, clusters) |> 
  summarize(
    mu = mean(wcss), 
    sigma = sd(wcss),
    .groups = 'drop'
  )

res2 =
  stats |>
  inner_join(
    kmeanres,
    by = c('trait','clusters')
  ) |>
  mutate(gap = log(mu) - log(wcss))

quants =
  res2 |>
  group_by(trait, null) |>
  slice_max(gap) |>
  group_by(trait) |>
  summarize(qtl = quantile(gap, .95))
```

```{r}
  
plot_gap = 
  res2 |>
  filter(null == 0) |>
  ggplot(aes(clusters, gap)) +
  geom_line() +
  geom_point() +
  geom_hline(
    aes(yintercept = qtl), 
    data = quants, 
    color = 'red'
  ) +
  facet_wrap(~trait)
  

plot_gap |>
  show()

```

